Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-01-22 manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ,man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 8206}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Best Practices in Latent Class Analysis using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}}
\date{}

\shorttitle{BEST PRACTICES LCA}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht}

\abstract{%
Latent class analysis is a family of techniques for identifying groups in data based
on a parametric model. Its examples include mixture models, latent profile
analysis, latent class analysis with ordinal indicators, and latent class
growth analysis. Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of latent class analysis,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture
models in the \texttt{tidySEM} R-package whose backend relies on
the \texttt{OpenMx} R-package.
}

@@

Latent class analysis (LCA) is an umbrella term
that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile analysis.
Vermunt, J.K. et al. (2004) defined LCA more generally as virtually any statistical model
where ``some of the parameters {[}\ldots{]} differ across unobserved subgroups''.

Despite its popularity, there is a lack of standards for estimating and reporting LCA.
While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed reporting guidelines
for a specific type of LCA known as latent growth models,
general reporting guidelines for all LCA are still lacking.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature
by suggesting updated guidelines for estimation and reporting of LCA,
based on current best practices.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples,
this paper exclusively relies on free, open source software in R.
Our goal is to make best-practices in LCA widely accessible.

@@

Latent class analysis is a group of methods for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of LCA can be understood in different ways.
A mixture model assumes that the study population
is composed of \(K\) subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate model is a normal distribution,
which can be described with two parameters: the mean and the variance.
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class,
and the probability that an individual belongs to each class.

@@

\caption{\label{tab:tabshoe}Observed group membership by estimated class membership.}
\centering
\begin{tabular}[t]{l|r|r}
\hline
Observed & Class 1 & Class 2\\
\hline
Man & 21 & 28\\
\hline
Woman & 51 & 0\\
\hline
\end{tabular}
\end{table}

A LCA is like confirmatory factor analysis,
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that
factor analysis can be considered as a way to group observed \emph{variables} into latent constructs,
and LCA groups \emph{individuals} into classes.
In line with this distinction, LCA is sometimes referred to as a ``person-centered'' technique,
and factor analysis as a ``variable-centered'' technique (Nylund-Gibson \& Choi, 2018).

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm.
In line with this perspective, LCA is sometimes described as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast ``model-based'' clustering refers to the fact that LCA
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, LCA can be considered as an
\emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
The term \emph{unsupervised} refers to the fact that the outcome variable
-- true class membership -- is not known,
and the term \emph{classification} refers to the fact that
the algorithm is predicting a categorical outcome -- class membership.

@@

In this paper, we use the term latent class analysis to refer to
a family of techniques that estimate latent class membership
based on a parametric model of observed indicators.
From a historical perspective, the term was initially conceived to
refer to analyses with categorical, usually binary indicators (Vermunt, J.K. et al., 2004).
Nowadays, there are a number of related techniques,
known by distinct names, that serve a similar purpose.
The term ``latent class analysis'' seems most appropriate
as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis,
and does not imply restrictions to the model used,
or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models,
we will provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the \emph{finite Gaussian mixture model};
a univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
The parameters of a finite Gaussian mixture model are
the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivariate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances
between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for instance when the cohesion between indicators is expected to differ between classes.
As an example, a LPA of ocean plastic particles
found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width,
meaning that these particles were approximately round or square in shape,
and a class of larger particles with a low correlation between length and width,
meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class,
one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated
based on these continuous latent variables.
A common application of this approach is in longitudinal research,
where the indicators reflect one construct assessed at different time points.
Examples of this approach include \emph{growth mixture models} (GMM)
and \emph{latent class growth analyses} (LCGA).
These techniques estimate a latent growth model
to describe individual trajectories over time.
The growth mixture model is a latent class model
where the parameters that indicate class membership
are the intercepts and variances,
and typically covariances of the latent growth variables,
e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero,
it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class
share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories
is due to residual error variance.

The term latent class analysis originally referred to cases
where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case
where indicators are of binary or ordinal measurement level
might be described as \emph{latent class analysis with ordinal indicators}.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that
each categorical variable reflects an underlying standard normal distribution.
The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution
(with \(N(\mu = 0, \sigma = 1)\)).
These thresholds are estimated based on
the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to all LCA methods.

@@

There are several use cases for which LCA methods are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time,
adolescents reside in one of four identity statuses.
Latent class analysis can be used to identify these four statuses
based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators,
then the theory may be called into question.

Another use case is unsupervised learning:
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid
when several clinical indicators can be used to distinguish between
a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
The example of shoe size is a rudimentary illustration of this type of application.

LCA can be used as a descriptive analysis
where a researcher wishes to describe their sample
and identify a few prototypes based on many variables.
As an example, if a survey among all Dutch academics was carried out
with the goal of bringing about a funding reform,
LCA could be used to discover the types of publications that get funding.

With LCA, our goal could be to inductively identify the number of classes.
For instance, if we believe that a variable represents a group,
but don't know how many groups there are,
LCA may be an appropriate technique to answer this question.
For example, Hopfer, Tan, and Wylie (2014) studied substance use, sexual behavior,
and mental health status of urban population in Winnipeg, Canada.
The underlying assumption was that there were different risk profiles,
but their number was not known.
From a collection of indicators, LCA provided evidence
that there are four distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals.
In a peer harassment study,
Giang and Graham (2008) used LCA to classify over 2,000 sixth grade students
into aggressor and victim latent classes.
The five-class solution comprised of classes of
victims, aggressors, and socially adjusted students.
For instance, it revealed that there were two types of victims:
highly-victimized aggressive-victims and highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators will be strongly related to the latent variable
and will lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on LCA, see Geiser and Wurpts (2014).
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ,man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 9001}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Best Practices in Latent Class Analysis using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}, Mauricio Garnier-Villarreal\textsuperscript{3}, \& Daniel Anadria\textsuperscript{1}}
\date{}

\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht\\\textsuperscript{3} Vrije Universiteit Amsterdam, Sociology}

\abstract{%
Latent class analysis (LCA) refers to a family of techniques for identifying groups
in data based on a parametric model.
Examples include mixture models, LCA with ordinal
indicators, and latent class growth analysis.
Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of LCA,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture
models in the \texttt{tidySEM} R-package whose backend relies on \texttt{OpenMx}.
}

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. Vermunt et al. (2004) defined LCA more generally as
virtually any statistical model where ``some of the parameters {[}\ldots{]}
differ across unobserved subgroups''.

Despite the popularity of LCA,
two challenges impede broader adoption and correct application of the method.
The first challenge is that most existing software that implements these models is commercial and closed-source.
The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
based on current best practice.
Furthermore, meeting the increasing demand for open source research software,
this paper introduces new functionality in the R-package \texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free open source software (FOSS).

\hypertarget{the-argument-for-free-open-source-software}{%
\subsection{The Argument for Free Open Source Software}\label{the-argument-for-free-open-source-software}}

Despite the popularity of LCA,
most existing software is commercial and closed-source,
for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which the R-package \texttt{MplusAutomation} offers augmented LCA functionality (\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD {[}REF{]}.
A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
This goes against open science principles (\textbf{lamprechtFAIRPrinciplesResearch2019?}),
incurs a risk that mistakes in the source code go unnoticed,
and curbs progress as independent developers cannot fix bugs or add functionality.

There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

Notable FOSS for LCA includes Mclust {[}REF{]}, depmixS4 {[}REF{]}, and OpenMx (Neale et al., 2016).
A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
or implement specific innovations (e.g., depmixS4).
OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
Remaining obstacles to the use of \texttt{OpenMx} is that it is not very user-friendly and its functionality for LCA is poorly documented.

These limitations are addressed by the \texttt{tidySEM} package,
which provide a user-friendly interface for LCA in \texttt{OpenMx}.
The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
but the resulting models can still be fully customized (low barrier, high ceiling).
Specifically, LCA models constructed in \texttt{tidySEM} inherit from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream functions that process such models.
Being FOSS, \texttt{tidySEM} facilitates fully-reproducible analyses and contributes to open science.
It was developed with open science in mind,
and adheres to the FAIR software principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}),
and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best Practices} for FOSS software.

@@

LCA refers to methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership (\textbf{vermuntLatentClassAnalysis2004?}).
It has become popular across scientific fields,
under a number of different names;
most notably (finite Gaussian) mixture modeling and latent profile analysis.
The concept of LCA can be understood in different ways.
Generally speaking, LCA assumes that the study population is composed of \(K\)
subpopulations.
It further assumes that the observed data are
a mixture of data originating from those subpopulations;
hence the name mixture model.
Consider the simplest possible
univariate ``model'', which is a normal distribution.
This model has
two parameters: the mean and the variance.
LCA aims to estimate the values of those parameters across a known number of classes \(K\),
as well as the probability of every individual \(i\) belonging to classes \(1 \ldots k\).
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).

@@

\caption{\label{tab:tabshoe}Observed group membership by estimated class membership.}
\centering
\begin{tabular}[t]{l|r|r}
\hline
Observed & Class 1 & Class 2\\
\hline
Man & 28 & 21\\
\hline
Woman & 0 & 51\\
\hline
\end{tabular}
\end{table}

Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}).
A mixture model is like confirmatory factor analysis,
except with a categorical instead of a continuous latent variable.
One difference between
the two techniques is that factor analysis can be considered as a way to
group observed \emph{variables} into latent constructs, and LCA groups
\emph{individuals} into classes.
While factor analysis seeks to describe the relations between variables,
LCA seeks to describe unobserved groups and classify individual observations to those groups.
In line with this distinction, LCA
(mixture models) is referred to as a ``person-centered'' technique, and
factor analysis as a ``variable-centered'' technique
(Masyn, 2013; Nylund-Gibson \& Choi, 2018).

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or individuals can be assigned to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
This perspective on mixture modeling is sometimes referred to as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast ``model-based'' clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

Finally, in the context of machine learning, LCA can be considered as an
\emph{unsupervised classification} method (Figueiredo \& Jain, 2002).
The term \emph{unsupervised} refers to the fact that the outcome variable --
true class membership -- is not known, and the term \emph{classification}
refers to the fact that the algorithm is predicting a categorical
outcome: class membership.

@@

The term LCA was initially conceived to refer to analyses with categorical,
usually binary indicators (Collins \& Lanza, 2009).
It was later generalized to refer to any technique that estimates a categorical latent variable based on a
parametric model of observed indicators, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators (Vermunt et al., 2004).
To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
Given the abundance of redundant terms in use for closely related types of LCA models,
we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated
freely, or set to be constrained across classes (Collins \& Lanza, 2009).

The technique known as \emph{latent profile analysis (LPA)} is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for,
the residual covariances/correlations between indicators are
assumed to be zero.
In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes (Collins \& Lanza, 2009).
Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
A class of larger particles with a low correlation between length and width,
and a class of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape,
whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).

Finally, some LCA models use latent variables as indicators.
A common application of this approach is in
longitudinal research with repeated measures of a construct.
Examples of this approach include
\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
(Jung \& Wickrama, 2008).
These techniques estimate a latent growth model
to describe grouped trajectories over time.
The indicators of class membership in a GMM are the
intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
For linear trajectories, these latent variables are an intercept and slope.
The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
This is achieved by restricting the (co)variances of the growth parameters to zero.
Any variance in the indicators not explained by the class-specific
latent trajectories is assumed to be error variance.

@@

There are several use cases for which LCA methods are suitable.

\hypertarget{testing-theory}{%
\subsubsection{Testing theory}\label{testing-theory}}

Although LCA is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that LCA is similar to confirmatory factor analysis,
it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) (\textbf{luyckxDevelopmentalTypologiesIdentity2008?}).
Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA (\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
A recent study by Maene, D'hondt, Van Lissa, Thijs, and Stevens (2022) applied this principle to
test the theory that identities of Belgian adolescents with migration background could be summarized as those relating to their heritage, national and regional identity.

\hypertarget{measurement-model}{%
\subsubsection{Measurement model}\label{measurement-model}}

LCA can also be used as a measurement model, similar to confirmatory factor analysis.
This is useful when class membership is measured imperfectly by several indicators:
LCA can partial out measurement error and restore most likely class membership.
For example, a recent study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
A mixture model was used to estimate ``error corrected employment status'', using self-report and registry data as indicators.
Among the outcomes of interest were random and systematic measurement error in both indicators,
which reveal strengths and weaknesses of both types of assessment.
